---
title: "Variable Selection"
output: html_notebook
runtime: shiny
editor_options: 
  markdown: 
    wrap: sentence
  fig_width: 10 
  fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(leaps)
library(corrplot)
library(dplyr)
```

```{r}
#setwd("Variable_Selection")
getwd()
```

## Reading the matrices

```{r}
#Rabies
transition_distances_rabies_mp<-read.csv("input/transition_distances_batRABV.MCC.keep.target.heights.treesMP.csv",  
                                         row.names = 1)
transition_distances_rabies_ml<-read.csv("input/transition_distances_batRABV.MCC.keep.target.heights.treesML.csv",  
                                         row.names = 1)
transition_distances_rabies_tt<-read.csv("input/transition_distances_batRABV.MCC.keep.target.heights.treesTT.csv", 
                                         row.names=1)

#Ebola
transition_distances_ebola_mp<-read.csv("input/transition_distances_ebola_not_annotated.treeMP.csv",  
                                        row.names = 1)
transition_distances_ebola_ml<-read.csv("input/transition_distances_ebola_not_annotated.treeML.csv", 
                                        row.names = 1)
transition_distances_ebola_tt<-read.csv("input/transition_distances_ebola_not_annotated.treeTT.csv",
                                        row.names = 1)

#Test
transition_distances_pdvc_mp<-read.csv("input/transition_distances_PDCV_discrete_MCC.treeMP.csv",
                                       row.names = 1)
transition_distances_pdvc_ml<-read.csv("input/transition_distances_PDCV_discrete_MCC.treeML.csv",
                                       row.names = 1)
transition_distances_pdvc_tt<-read.csv("input/transition_distances_PDCV_discrete_MCC.treeTT.csv", 
                                       row.names = 1)

#Influenza
transition_distances_influenza_mp<-read.csv("input/transition_distances_h3_small_sample.MCC.treMP.csv", 
                                            row.names = 1)
transition_distances_influenza_ml<-read.csv("input/transition_distances_h3_small_sample.MCC.treML.csv", 
                                            row.names = 1)
transition_distances_influenza_tt<-read.csv("input/transition_distances_h3_small_sample.MCC.treTT.csv", 
                                           row.names = 1)

transition_distances_rabies_mp <- transition_distances_rabies_mp[colnames(transition_distances_rabies_mp)!="Key"]
transition_distances_rabies_ml <- transition_distances_rabies_ml[!colnames(transition_distances_rabies_ml) %in% c("Transition_Rates", "Key")]
transition_distances_rabies_tt <- transition_distances_rabies_tt[colnames(transition_distances_rabies_tt)!="Key"]
transition_distances_ebola_mp <- transition_distances_ebola_mp[colnames(transition_distances_ebola_mp)!="Key"]
transition_distances_ebola_ml <- transition_distances_ebola_ml[!colnames(transition_distances_ebola_ml) %in% c("Transition_Rates", "Key")]
transition_distances_ebola_tt <- transition_distances_ebola_tt[colnames(transition_distances_ebola_tt)!="Key"]
transition_distances_pdvc_mp <- transition_distances_pdvc_mp[colnames(transition_distances_pdvc_mp)!="Key"]
transition_distances_pdvc_ml <- transition_distances_pdvc_ml[!colnames(transition_distances_pdvc_ml) %in% c("Transition_Rates", "Key")]
transition_distances_pdvc_tt <- transition_distances_pdvc_tt[colnames(transition_distances_pdvc_tt)!="Key"]
transition_distances_influenza_mp <- transition_distances_influenza_mp[colnames(transition_distances_influenza_mp)!="Key"]
transition_distances_influenza_ml <- transition_distances_influenza_ml[!colnames(transition_distances_influenza_ml) %in% c("Transition_Rates", "Key")]
transition_distances_influenza_tt <- transition_distances_influenza_tt[colnames(transition_distances_influenza_tt)!="Key"]
```

```{r}
matrices_list <- list(
  "transition_distances_rabies_mp"= transition_distances_rabies_mp ,
   "transition_distances_rabies_ml" = transition_distances_rabies_ml ,
   "transition_distances_rabies_tt" = transition_distances_rabies_tt ,
   "transition_distances_ebola_mp" = transition_distances_ebola_mp ,
   "transition_distances_ebola_ml" = transition_distances_ebola_ml ,
   "transition_distances_ebola_tt" = transition_distances_ebola_tt ,
   "transition_distances_pdvc_mp" = transition_distances_pdvc_mp ,
   "transition_distances_pdvc_ml" = transition_distances_pdvc_ml ,
   "transition_distances_pdvc_tt" = transition_distances_pdvc_tt ,
   "transition_distances_influenza_mp" = transition_distances_influenza_mp ,
   "transition_distances_influenza_ml" = transition_distances_influenza_ml ,
   "transition_distances_influenza_tt" = transition_distances_influenza_tt 
)
```

## Choose matrix to analyze
```{r}
selectInput("matrix",
            label = "Transition and Distances Matrix",
            choices =  names(matrices_list))
```

```{r}

output$response <-renderUI({
  req(input$matrix)
  selectInput("response",
            label = "Response variable",
            choices= colnames(matrices_list[[input$matrix]] %>% dplyr::select(starts_with("Transition"))))
    })

uiOutput("response")
```

## Correlation Plot
```{r, fig.width = 20, fig.height = 20}
renderPlot({
  cor_matrix<-cor(matrices_list[[input$matrix]])
  corrplot::corrplot(cor_matrix, type="upper", order="hclust", title = input$matrix, tl.cex = 1.5)
},
width = 800, 
height = 800)
```

## Ordinary Least Squares Regression

```{r}
lm_ols<-reactive({
  req(input$response, input$matrix)
  f<-as.formula(paste0(input$response, "~",
                       paste(colnames(matrices_list[[input$matrix]]%>%select(!starts_with("Trans"))), collapse = "+")))
  
  lm <- lm( formula=f, data=matrices_list[[input$matrix]])

  lm[["call"]][["formula"]]<-eval(lm[["call"]][["formula"]])
  lm[["call"]][["data"]]<-eval(input$matrix)
  lm
})

renderPrint({
  summary(lm_ols())
})

```

## Leaps package : Best model for every amount of predictors, forward and backward selection

### Select the method
```{r}
selectInput("step_method",
            label = "Stepwise selection method:",
            choices =  c("exhaustive", "backward", "forward"))
```

The variable to be included is marked with an asterix:
```{r}
lm_regsubsets<-reactive({
  req(input$response, input$matrix)
  lm_regsubsets <- leaps::regsubsets(
     x=as.formula(paste0(input$response, "~", 
                       paste(colnames(matrices_list[[input$matrix]]%>%select(!starts_with("Trans"))), collapse = "+"))),
     data=matrices_list[[input$matrix]],
     nvmax = min(length(colnames(matrices_list[[input$matrix]]))),
     method=input$step_method)
    
  lm_regsubsets[["call"]][[2]]<-as.formula(paste0(input$response, "~", paste(colnames(matrices_list[[input$matrix]]%>%select(!starts_with("Trans"))), collapse = "+")))
 # lm_regsubsets[["call"]][[3]]<-eval(input$matrix) 
  lm_regsubsets
})

renderPrint({
  summary(lm_regsubsets())
})
```

# 
# ```{r}
# library(dplyr)
# lm_regsubsets2 <- leaps::regsubsets(
#      x=as.formula(paste0("Transitions", "~",
#                        paste(colnames(matrices_list[["transition_distances_influenza_tt"]]%>%select(!starts_with("Trans"))), collapse = "+"))),
#      data=matrices_list[["transition_distances_influenza_tt"]],
#      nvmax = min(10, length(colnames(matrices_list[["transition_distances_influenza_tt"]]))),
#      method="forward")
# sum_regsubsets2<-summary(lm_regsubsets2)
# sum_regsubsets2
# names(sum_regsubsets2)
# ```
# 

### Select the criterium
```{r}
selectInput("selection_crit",
            label = "Selection criterium:",
            choices = c("rsq", "rss", "adjr2", "cp", "bic"))
```

```{r}
renderPlot({
  sum_regsubsets<-summary(lm_regsubsets())
  plot(get(input$selection_crit, sum_regsubsets), xlab="Number of variables", ylab=input$selection_crit)
})
```

BIC is  more stringent then the other, also you see how RSS and R² are not valid selection criteriums since they strictly improve with additionaly predictors.


### What is the best number of selected variables (also check plots)
```{r}
renderPrint({
  if(input$selection_crit %in% c("cp", "bic", "rss")){
    which(summary(lm_regsubsets())[[input$selection_crit]]==min(summary(lm_regsubsets())[[input$selection_crit]]))
  }else{
    which(summary(lm_regsubsets())[[input$selection_crit]]==max(summary(lm_regsubsets())[[input$selection_crit]]))
  }
})
```

```{r}
output$coefs<-renderUI({
  numericInput(
    inputId= "number_of_vars",
    label="Get coefficients for selected number of var:",
    value=which(summary(lm_regsubsets())[[input$selection_crit]]==max(summary(lm_regsubsets())[[input$selection_crit]])),
    min=1)
})
uiOutput("coefs")
```

```{r}
renderPrint({
  req(input$number_of_vars)
  coef(lm_regsubsets(), input$number_of_vars)
})
```



<!-- ### Lasso selection -->

<!-- ```{r} -->
<!-- library(glmnet) -->
<!-- grid<-10^seq(10,-2, length=100) #100 values of lambda covering the range 0.01 to 10^10 -->
<!-- x<-model.matrix(Transitions~., transition_distances[,2:27])[,-1] #this means everything except column 1, the intercept -->
<!-- y<-transition_distances$Transitions -->
<!-- set.seed(1) -->
<!-- train<-sample(1:nrow(x), nrow(x)/2) -->
<!-- test<-(-train) -->
<!-- y.test<-y[test] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- set.seed(1) -->
<!-- cv.out<-cv.glmnet(x[train,], y[train], alpha=1, type.measure = "deviance", nfolds = 10) #deviance a.k.a MSE -->
<!-- plot(cv.out) -->
<!-- ``` -->

<!-- Best tuning parameter we can select based on the mean squared error/deviance as we also see in the plot above: -->

<!-- ```{r} -->
<!-- bestlam<-cv.out$lambda.min -->
<!-- bestlam -->
<!-- ``` -->

<!-- Assessing how good the model is by predicting the number of transitions for the non-used test observations and then calculating the MSE with the real transitions. -->

<!-- ```{r} -->
<!-- lasso.pred<-predict(cv.out, s=bestlam, newx=x[test,]) -->
<!-- mean((lasso.pred-y[test])^2) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- set.seed(1) -->
<!-- out<-glmnet(x,y,alpha=1, lambda=0) -->
<!-- lasso.coef<-predict(out, s=bestlam, type="coefficients")#[1:25] -->
<!-- lasso.coef -->
<!-- ``` -->

<!-- ```{r} -->
<!-- pred<-predict(out,s=0, newx=x) -->
<!-- cor.test(pred,y) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- summary(lm(Transitions ~ . , data =transition_distances[,2:27]))$coefficients -->
<!-- ``` -->

<!-- ```{r} -->
<!-- names(data) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- library(plotmo) -->
<!-- plot_glmnet(out, xvar="lambda", xlim=c(-3,1), ylim=c(-2,2), col=1:25, label=T) -->
<!-- abline(v=log(bestlam), lty=2) -->
<!-- ``` -->

<!-- The model has only 4 predictors left and consistently greatCircleDistances is among the selected predictors. -->
